Bayesian Rate Estimation

### 2012-08-16 many.tex
Updates to analytic example per Ilya's comments; better figures, more captions.

For the usual $k=1$ case, the marginalised posterior
(Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{equation}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right)\exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
where $x_N$ denotes the loudness of the loudest event. Marginalising
over $R_b$ we obtain
\begin{equation}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left(
\frac{\hat{b}(x_N,\theta)}{2(1-\hat{B}(x_N,\theta))} + R_f
\hat{f}(x_N,\theta)\right)\frac{\sqrt{\pi}}{\sqrt{1-\hat{B}(x_N,\theta)}\,\sqrt{R_f}}
\exp(-R_f(1-\hat{F}(x_N,\theta))) .
\end{equation}
This posterior has a maximum in $R_f$ at
\begin{equation}
R_f =
\frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta)\tilde{b}(x_N,\theta)
  \pm
  \sqrt{(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta)\tilde{b}(x_N,\theta))^2
    -
    4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)}}{4
  \hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta)}.
\end{equation}
where $\tilde{b}(x_N,\theta)
=\hat{b}(x_N,\theta)/(1-\hat{B}(x_N,\theta))$. If
$\tilde{b}(x_N,\theta) \ll 1$, i.e., the loudest event is very likely
to be foreground, we obtain the result $(1- \hat{F}(x_N,\theta))\,R_f
\approx 1/2$. This is can be understood as the statement that the rate
of foreground events with ranking statistic greater than $x_N$, $(1-
\hat{F}(x_N,\theta))\,R_f$, is of order $1$, as expected. If
$\tilde{b}(x_N,\theta) >> 1$, however, the posterior on $R_f$ is
peaked at $0$.

@@

\ilya{[General comment on the text: I think we need to say a bit more
    about the figures and the interpretation of the results, instead
    of just relying on the reader to make the best of it.  For
    example, even though it may seem obvious to us, we should
    explicitly say that Fig.~2 shows that we succeed in recovering
    rates within the expected uncertainty envelope.  On Fig.~3, could
    say something obvious like point out that we cannot separate
    background from foreground for individual events, but that doesn't
    necessarily matter for rate measurements.  Now, one interesting
    question that Fig.~3 brings to mind is the following: if we only
    used "gold-plated" events (say, those with
    $p(x|fore)/p(x|back)>0.99$, how much worse / more uncertain would
    the rate estimation be?]}

\begin{figure}
  \includegraphics[width=\columnwidth]{rates}
  \caption{\label{fig:analytic-rate-recovery} The marginalized
    posterior densities for $R_f$ (solid line) and $R_b$ (dashed line)
    for the analytic model discussed in \S
    \ref{sec:analytic-GW-example}.  The vertical lines indicate the
    ``true'' values used to generate the synthetic data set.}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{pfore}
  \caption{\label{fig:analytic-rate-foreground-probs} Foreground
    probability for each event in the synthetic data set of \S
    \ref{sec:analytic-GW-example} marginalized over all other
    parameters.  True foreground events are in blue \ilya{(dark
      grey)}, background events in red \ilya{(light grey)}.  The solid
    line is the likelihood ratio
    $p(x|\mathrm{foreground})/p(x|\mathrm{background}) =
    \hat{f}(x)/\hat{b}(x)$ for this model \ilya{evaluated at the
      simulated parameters (?)}; this exceeds the marginalized
    foreground probability for many of the events because in this data
    set there are approximately nine times as many background events
    as foreground events. \ilya{[I'm not sure what the reader is meant
        to take away from the solid lines If there's something worth
        taking away, we should describe the message in the text.
        Otherwise, if it only confuses things, we should drop it.]}}
\end{figure}

@@

In \S \ref{sec:analytic-GW-example} we assumed that the overlap
between \ilya{different} templates in the template bank \ilya{was
  negligible}, so the SNRs \ilya{recovered by} different templates are
independent random variables.  In fact, template banks are not
constructed in this way \citep[e.g.,][]{Owen:1998dk,Ajith:2008},
%\citep{Caudill2011,Cannon2011}, 
because signals could fall in the gaps between the non-overlapping
templates.  We can model this effect by assuming that a template bank
of $N$ actual templates will behave as if it had $N_\mathrm{eff}$
\emph{independent} templates.  \ilya{Rather than pre-computing
  $N_\mathrm{eff}$,} we can fit for \ilya{it as a} shape parameter.
That is, we assume that \ilya{$\theta = \{N_\mathrm{eff}\}$} is a
shape parameter for the background cumulative distribution:
\begin{equation}
  \hat{B}\left(x, N_\mathrm{eff}\right) = \frac{\left( 1 + \erf\left(
    \frac{x}{\sqrt{2}} \right) \right)^{N_\mathrm{eff}} - \left( 1 +
    \erf\left( \frac{\xmin}{\sqrt{2}} \right)
    \right)^{N_\mathrm{eff}}}{2^{N_\mathrm{eff}} - \left( 1 +
    \erf\left( \frac{\xmin}{\sqrt{2}} \right) \right)^{N_\mathrm{eff}} }.
\end{equation}

Results from such an analysis appear in Figures \ref{fig:rates-nt} and
\ref{fig:ntemplates}.  We use the same parameters and data set as in
\S \ref{sec:analytic-GW-example}, with $R_f = 10.4$, $R_b = 95.1$, and
$N_\mathrm{eff} = 1000$, but now allow $N_\mathrm{eff}$ to be a
parameter of the background distribution, with a flat prior.  Both the
rates and the number of effective templates are recovered without
significant loss of accuracy relative to the fixed $N_\mathrm{eff}$
situation in \S \ref{sec:analytic-GW-example}.

>>>>

For the usual $k=1$ case \citep{Biswas2009}, the marginalised
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{equation}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right)\exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
where $x_N$ denotes the loudness of the loudest event. Marginalising
over $R_b$ we obtain
\begin{equation}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left(
\frac{\hat{b}(x_N,\theta)}{2(1-\hat{B}(x_N,\theta))} + R_f
\hat{f}(x_N,\theta)\right)\frac{\sqrt{\pi}}{\sqrt{1-\hat{B}(x_N,\theta)}\,\sqrt{R_f}}
\exp(-R_f(1-\hat{F}(x_N,\theta))) .
\end{equation}
This posterior has a maximum in $R_f$ at
\begin{equation}
R_f =
\frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta)\tilde{b}(x_N,\theta)
  \pm
  \sqrt{(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta)\tilde{b}(x_N,\theta))^2
    -
    4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)}}{4
  \hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta)}.
\end{equation}
where $\tilde{b}(x_N,\theta)
=\hat{b}(x_N,\theta)/(1-\hat{B}(x_N,\theta))$. If
$\tilde{b}(x_N,\theta) \ll 1$, i.e., the loudest event is very likely
to be foreground, we obtain the result $(1- \hat{F}(x_N,\theta))\,R_f
\approx 1/2$. This is can be understood as the statement that the rate
of foreground events with ranking statistic greater than $x_N$, $(1-
\hat{F}(x_N,\theta))\,R_f$, is of order $1$, as expected. If
$\tilde{b}(x_N,\theta) >> 1$, however, the posterior on $R_f$ is
peaked at $0$.

@@

\ilya{Now, one interesting question that Fig.~3 brings to mind is the
  following: if we only used "gold-plated" events (say, those with
  $p(x|fore)/p(x|back)>0.99$, how much worse / more uncertain would
  the rate estimation be?]}

\begin{figure}
  \includegraphics[width=\columnwidth]{rates}
  \caption{\label{fig:analytic-rate-recovery} The marginalized
    posterior densities for $R_f$ (solid line) and $R_b$ (dashed line)
    for the analytic model discussed in \S
    \ref{sec:analytic-GW-example}.  The vertical lines indicate the
    ``true'' values used to generate the synthetic data set.  Both the
    true foreground and background rates lie well within the
    probability envelope for $R_f$ and $R_b$.}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{pfore}
  \caption{\label{fig:analytic-rate-foreground-probs} Foreground
    probability for each event in the synthetic data set of \S
    \ref{sec:analytic-GW-example} marginalized over all other
    parameters (errorbars are one standard deviation).  True
    foreground events are in dark grey, background events in light
    grey.  Even though our method cannot identify the status of most
    events with confidence, it can still correctly estimate the rates
    (Figure \ref{fig:analytic-rate-recovery}).}
\end{figure}

@@

In \S~\ref{sec:analytic-GW-example} we assumed that the overlap
between different templates in the template bank was negligible, so
the SNRs recovered by different templates are independent random
variables.  In fact, template banks are not constructed in this way
\citep[e.g.,][]{Owen:1998dk,Ajith:2008}, because signals could fall in
the gaps between the non-overlapping templates.  We can model this
effect by assuming that a template bank of $N$ actual templates will
behave as if it had $N_\mathrm{eff}$ \emph{independent} templates.
Rather than pre-computing $N_\mathrm{eff}$, we can fit for it as a
shape parameter.  That is, we assume that $\theta =
\{N_\mathrm{eff}\}$ is a shape parameter for the background cumulative
distribution:
\begin{equation}
  \hat{B}\left(x, N_\mathrm{eff}\right) = \frac{\left( 1 + \erf\left(
    \frac{x}{\sqrt{2}} \right) \right)^{N_\mathrm{eff}} - \left( 1 +
    \erf\left( \frac{\xmin}{\sqrt{2}} \right)
    \right)^{N_\mathrm{eff}}}{2^{N_\mathrm{eff}} - \left( 1 +
    \erf\left( \frac{\xmin}{\sqrt{2}} \right) \right)^{N_\mathrm{eff}} }.
\end{equation}

Results from such an analysis on the data in
\S~\ref{sec:analytic-GW-example} appear in Figures \ref{fig:rates-nt}
and \ref{fig:ntemplates}.  Recall that the parameters used to generate
this data set are $R_f = 10.4$, $R_b = 95.1$, and $N_\mathrm{eff} =
1000$.  We use a flat prior on $N_\mathrm{eff}$.  Both the rates and
the number of effective templates are recovered without significant
loss of accuracy relative to the fixed $N_\mathrm{eff}$ situation in
\S~\ref{sec:analytic-GW-example}.
